---
title: "sortingmod"
author: "Thomas de Graaff and Or Levkovich"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sortingmod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
`sortingmod` is a package for estimating the sorting model - a discrete choice model which explains the location decision of heterogeneous individuals over a set of alternative locations. The model is developed by Bayer et al. (2004) following the work of Berry et al. (1995). It relies on the assumptions that individuals choose a location that maximizes their utility, and that heterogeneous individuals with different characteristics have different preferences, and different valuation for location characteristics. The results of the model provide choice probabilities for each alternative, as well as insight on the valuation patterns of heterogeneous agents, and marginal willingness-to-pay values for location characteristics.

The package is constructed as an accompanying tool for the Summer School in “Hedonic price analysis and the residential location choice”, hosted by the Kraks Fond - Institute for Urban Economic Research (Copenhagen, Denmark).

## Installation



## Estimation of the residential sorting model
`Sortingmod` package follows the methodology described in Bayer et al. (2004) to analyze how heterogeneous individuals sort themselves into alternative locations based on their preferneces for local characteristics. 

`sortingmod` solves the following model:

$$U_{j}^{i}  =  \sum_{k=1}^{K}\beta_{k,0}X_{k,j} + \sum_{k=1}^{K}\sum_{l=1}^{L}\beta_{k,l}(Z^{i}_{l}-\bar{Z_{l}})X_{k,j} + \xi_j + \epsilon_{j}^{i}$$

Where:
		$U_{j}^{i}$ is the utility of individual $i$ from alternative $j$,
		$X_{k,j}$ is the value of the $k$-th alternative characteristic in location $j$, 
		$Z^{i}_{l}$ is the value of the $l$-th characteristic of individual $i$,
		$\bar{Z_{l}}$ is the sample mean of individual characteristic $l$,
		$\beta_{k,l}$ is the cross effect coefficients,
		$\xi_j$ is an alternative specific error term, and
		$\epsilon_{j}^{i}$ is the individual error term.  
		
Due to likely correlation between prices and unobserved characteristics, it is necessary to estimate the sorting model in two stages.(See discussion in Berry et al. (1995)).


### First stage
The first stage of the sorting model involves estimating the cross effect coefficients, and treating $\delta_{j} = \sum_{k=1}^{K}\beta_{k,0}X_{k,j}$ as alternative specific constants. Assuming that the random term is IID extreme value type I distributed, choice probabilities can be derived by a logit model estimation:

$$P^{i}_{j} = \frac{e^{V_{j}^{i}}}{\sum_{j=1}^{J}e^{V_{j}^{i}}} = \frac{e^{(\delta_{j} + \sum_{k=1}^{K}\sum_{l=1}^{L}\beta_{k,l}(Z^{i}_{l}-\bar{Z_{l}})X_{k,j})}}{\sum_{j=1}^{J}e^{(\delta_{j} + \sum_{k=1}^{K}\sum_{l=1}^{L}\beta_{k,l}(Z^{i}_{l}-\bar{Z_{l}})X_{k,j})}}$$

The cross-effect coefficients and alternative specific constants are estimated by maximizing the probability that individual $i$ choose location $j$, using a maximum likelihood procedure (BHHH (Berndt et al., 1974)).

### Second stage
The second stage of the estimation involves estimating the equation: 

$$\delta_{j} = \sum_{k=1}^{K}\beta_{k,0}X_{k,j} + \xi_j$$

where the vector of mean indirect utilities ($\delta_{j}$) is explained by the alternative characteristics. 


### Endogeneity in the observed alternative characteristics

Since prices are likely to be correlated with the error term $\xi_j$ the second stage estimation follows a two stage least squares (2SLS) procedure to account for this endogeneity problem.
Here, we follow the approach of \citet{Bayer2004} for instrument construction and construct a price instrument assuming no correlation exists with unobserved location heterogeneity. Namely, we compute the price vector which would clear the market under the assumption of $(\xi_j = 0)$.


## Quick start

The purpose of this section is to give a general sense of the package, its components and an overview of how to use it. 

First, we load the package :
```{r}
library(sortingmod)
```

### Data structure

The data for the estimation should be in data frame format, and it must contain: 

- One or more columns indicating characteristics of a unique agent making a discrete choice between available alternatives (individuals, households, etc.).
- A vector indicating the chosen alternative. 
- One or more columns indicating characteristics of the chosen alternative.

Users can work on their own data, or use the data which is saved in the workspace.

```{r, echo = FALSE}
load("~/sortingmod/data/municipality.rda")
code_name <- c("mun_code")
Z_names <- c("age","income","hh_kids")
X_names <- c("lnprice","nature","monuments")
head(municipality[,c("id",Z_names,code_name,X_names)])
```

Notes: 

- Each row represents a unique individual. The number of alternatives should therefore be smaller or equal to the number of individuals.
- Individual and alternative characteristics should be numeric variables. Factor variables are not yet supported, so it's best to transform them into numeric dummy variables if you intend to include them in the model.

### First stage estimation

The function `first_stage` is used in order to estimate the first stage of the model. `first_stage` takes four arguments: 

- `code_name` - which indicates (with name or column number) the vector with alternative chosen.
- `Z_names` - which indicates (with names or column numbers) the vectors with individual data.
- `X_names` - which indicates (with names or column numbers) the vectors with alternative data.
- `data` - which indicates the Data to be used.

```{r, results = "hide"}
code_name = c("mun_code")
X_names = c("lnprice","kindergardens_1km","p_mig_west","nature","monuments","cafes_1km")
Z_names = c("income","double_earner_hh","hh_kids","age", "migskill")
s1.results <- first_stage(code_name, Z_names, X_names, data = municipality)
```

s1.results is a (maxLik) object, containing the estimation results and all relevant information. Several methods can be called on the object, including `summary, coef,` and `stdEr`. Calling `summary` provides a neat summary of the estimation, including the coefficients and std. errors of the cross-effect coefficients and the alternative-specific constants, which will later be used in the second stage of the estiamtion. 


### Second stage estimation

The function `second_stage` is used in order to estimate the first stage of the model. `second_stage` takes four arguments: 

- `s1.results` - which indicates the (maxLik) object estimation results of the first stage of the sorting model.
- `data` - which indicates the Data to be used.
- `endog` - (optional) which indicates the endogenous variable(s) to be instrumented (from the dataset in parentheses).
- `instr` - (optional) which indicates the intrument(s) for the endogenous variable.


```{r, eval = FALSE}
second_stage(s1.results, data)

endog <- ("lnprice")
second_stage(s1.results, data = municipality, endog = endog, instr = instrument)
```

Before estimating the second stage of the model it is important to consider that some alternative characteristics are likely to be endogenous. For example, local price levels are very likely to be correlated with unobserved alternative features. For this reason the `second_stage` function allows (and encourages) users to include instruments in this part of the model estimation, using the options `endog` to indicate the endogenous variable(s), and `instrument` to indicate the chose instrument(s).

### Calculating the market equilibrium instrument

`sortingmod` also includes the function `sorting_inst`, which generates an instrument which rises "naturally" from the model and utilizes its equilibrium properties, following Bayer et al. (2004). 
The calculated instrument is an instrument which clears the market assuming no correlation exists with unobserved heterogeneity between alternatives, and it is calculated in an interative process based on the results of the first stage of the estimation process.

```{r, results = "hide"}
endog <- ("lnprice")
phat <- sorting_inst(s1.results, endog, data = municipality, stepsize = 0.02)
```

`sorting_inst` takes several (non-optional) arguments: 

- `s1.results` - which indicates the (maxLik) object estimation results of the first stage of the sorting model.
- `endog` - which indicates the endogenous variable to be instrumented.
- `data` Dataset to be used.

Unlike `first_stage`, `sorting_inst` does not take chosen alternatives, individual characteristics or alternative characteristics as arguments. This is because they are already specified as arguments in the `s1.results` object.
However, note that `data` must indicate the same data used for the first stage estimation, and that the indicated endogenous variable must be included in the set of alternative characteristics used in the estimation (and specified in `X_names`).

`sorting_inst` returns a list which contains (1) Results of the IV estimation, with the computed vector as instrument for the endogenous variable. (2) a vector of the computed instrument, and (3) the correlation between the computed instrument and the original variable.

```{r, echo = FALSE}
summary(phat$IV_results)
```

Notes: 

- The iteration process breaks if it identifies that values are not converging. Lack of convergence can be due to too much unobserved heterogeneity in the model, but it may also be that the contraction mapping coefficient (`stepstize` argument) is too large, which cause the iteration to overshoot. So before updating your model, try to experiment and choose lower stepsize values.
- `sorting_inst` accepts only one argument for endogenous variable as input, and subsequently also returns only a single vector as instrument. 


## Post estimation
### Marginal willingness-to-pay values
### References

Bayer, P., McMillan, R., & Rueben, K. (2004). An equilibrium model of sorting in an urban housing market (No. w10865). National Bureau of Economic Research.

Berry, S., Levinsohn, J., & Pakes, A. (1995). Automobile prices in market equilibrium. Econometrica: Journal of the Econometric Society, 841-890.

